---
title: 'SMSC Migratory Birds: Home Range Lab'
author: "N Simon Sanghera"
date: "13/12/2019"
output: html_document
---

#------------
Taught by Luke Powell  
  
**Overview**  
Learn to calculate animal home ranges using the adehabitatHR package in R.  
Types will be Minimum Convex Polygon and Kernel Density Estimator  
Also, how to calculate home ranges in 3D.  
  
  https://ecosystems.psu.edu/research/labs/walter-lab/manual
  https://scbi-migbirds.github.io/homeRange.html
  
#### Set-up  
\  

```{r, setup, include = FALSE}
library(knitr)
opts_knit$set(root.dir="F:/Science Projects/SMSC/SMSC_Migratory_Birds/")

```


Install packages and load libraries  
  
```{r libraries}
libraries <- 
  c(
    "tidyverse",
    "ggplot2",
    "adehabitatHR",
    "maptools"
  )

purrr::map(
  libraries,
  function(x) {
    library(x, character.only = TRUE)
  }
)



```
  
  
Import the birds location data. The data are in UTM format, units are metres.  
  
```{r import data}
locs <- read.csv("HomeRange/HR_data/RSHomeRangeData.csv",
                  header = T,
                  na.strings = c("NA", "NULL", "", "."))
# na.strings replaces missing values with "NA"

#locs <- import

```
  
  
Have a quick look at the data for each of the three birds  
  
```{r locs check}
table(locs$bird)

```
  
  
  
Import the telemetry data  
  
```{r telem}
telem <- read.csv("HomeRange/HR_data/RawBrazilTelemData26aug2014.csv",
                  header = T,
                  na.strings = c("NA", "NULL", "", "."))

```

  
### Prepare for Triangulations  
  
_Raw vs Adjusted Azimuth_ (compass bearing)  
It is possible to use an uncorrected compass and make corrections later, or use a compass that aloows adjustments to declination to be made on the compass itself.  
Or we can calculate declination (the angular distance of a point north or south of the celestial equator).  
At most places on the Earth's surface, the compass does't point exactly towards geographic north. The deviation ofthe compass from **TRUE NORTH** is an angle called **DECLINATION** (or magnetic declination).  
  
#### Conditions that affect radioteletry:  
  - topography 
  - wet vegetation  
  - distance to animal  
  - size of transmitter  
  - animal behaivour  
  
#### How to triangulate data  
  Two pieces of software are readily available on PCs:  
  - LOCATE III: free program, but buggy  
  - LOAS: expensive, not buggy  
  
#### Home-Range: how many points until home range stabilizes?  
  - 2D: usually 30 - 50 _independent_ points is usually enough (Seamen _et al._ 1999, JWM)  
  - 3D: around 80 independent points (Cooper _et al._, 2014, The Auk)  
  - Discuss biological (Lair, 1987) vs statistical independence  
  - It is possible to use bootstrapping to perform an asymptote analysis  
  - One can check the animals one-by-one (although this is time consuming)  
  
  
#### Note:  
Home Range Area Curve = plots the number of independent locations on the x-axis against the estimated home-range size on the y-axis. From the plot, the number of required locations is denoted when increasinf the number of locations does NOT result in an increasing home-range size (i.e. asymptotes of the graph) (Haines _et al._, 2009)  
  
  
### Prepare data for adehabitatHR  
  
Extracting just the x and y data  
```{r extract x and y data}
xyt <- subset(locs, select = c(X, Y))

```
  

Extracting just the bird IDs  
```{r extract bird IDs}
bird_id <- subset(locs, select = bird)

```

  

Creating a SpatialPointsDataFrame for the package adehabitatHR  
**SpatialPointsDataFrame** =  
  
```{r create spatialpointsdataframe}
locs_1              <- bird_id  
coordinates(locs_1) <- xyt

class(locs_1)

```
  
  
Plot the point data  
  
```{r plot initial point data}
plot(locs_1,
     col = as.data.frame(locs_1)[, 1])

```
  
  
### Minimum Convex Polygons  
  
Running the 95% Minimum Convex Polygon analysis  
This method calculates the smallest polygon enclosing 95% (removing 5% of the most extreme points) of the relocations of the animal. It is considered to be the home range of the animal.  
_home range_ is defined as the area traversed by the animal during its **normal** activities of foraging, mating, caring for young. Due to this reason, it is common to remove a small percentage of the relocations furthest from the centroid of the cloud of relocations before estimations.  
MCP can be used to describe the extent of distribution of locations of an animal but **NOT** as an estimation of home rnage size.  
  
```{r calc mcp 95}
locs_1_cp <- mcp(locs_1[, 1],
                 percent = 95)

```
  
  
```{r graph of polygons and points}
plot(locs_1_cp)

plot(locs_1,
     col = as.data.frame(locs_1)[, 1],
     add = T)
# adding colour for each species

```
  
  
Obtain the area of the polygon  
  
```{r area of polygon}
locs_1_cp

```
  
  
Write the polygon to a shapefile  
```{r write shapefile}
writePolyShape(locs_1_cp, "HomeRange/HR_data/RS_mcp")

```
  
writePolyShape is deprecated; use rgdal::writeOGR or sf::st_write  
  
  
  
### Kernel Density Estimation in 2D  

#==== Kernel Density Estimation ====#
This is a non-parametric method of estimating the probability density function of a random variable. It is a fundamental data smoothing problem where inferences about the population (data points/locations) are made based on a finited data sample.  
It is used to delineate where geo-tracked animals spend their time.  
KDE optimizers are sensitive to temporal autocorrelation.  
#==== ====#
  
Now we will calculate home range size using the kernel density estimator.  
It uses a _Utilization Distribution_ (UD) to calculate home range size based on the distribution of the points.  
Under this model, it is considered that animals use of space can be described by a _bivariate_ probability density function.  
This gives the probability density to relocate the animal at any place according to the coordinates (x, y) of this place.  
Home range sizes can be calculated by running the _Kernel Density Estimation_ (KDE) using the _Least Square Cross Validation_ technique.  
A bivariate kernel function is placed over each relocation, and the values of these functions are averaged together.  
Least-Squares Cross Validation is suggested to be better at identifying distributions with tight clumps.  
(Note: h is the smoothing parameter - this controls the width of the kernel functions placed over each point)  
  
```{r obtain kernel UD}
locs_1_kud <- kernelUD(locs_1[, 1],
                       h = "LSCV")

```
  
  
Converting the KDE as vectors:  
```{r kde vectors}
homerange_locs_1 <- getverticeshr(locs_1_kud)


```
  
  
```{r class of vectors}
class(homerange_locs_1)

```
  
  
Plot the vector  
```{r plotting the vector}
plot(homerange_locs_1,
     col = 1:3)

```
  
  
Calculating the area for the different _isopleths_ (UDs).  
- areas are in hectares  
- 95% is the standard for "home range size"  
- 50% is often used for the core area  
  
```{r calculate area of kd}
locs_1_kde_area <- kernel.area(locs_1_kud,
                               percent = c(50, 95))

```
  
  
```{r kde area}
locs_1_kde_area

```
  
  
Plot the different isopleths:  
- for the second animal only  
  
```{r plot isopleth for animal 2}
# First get volume of the UD
vud_2 <- getvolumeUD(locs_1_kud)
image(vud_2[[2]])
xyzv <- as.image.SpatialGridDataFrame(vud_2[[2]])
contour(xyzv, add = TRUE)

```
  
  
GPS data are autocorrelated - so they can pose difficulties in estimating the bandwidth and violate assumptions of independence of locations that are inherent to the KDE.  
Datasets for avian and mammal species can include > 10,000 locations, and only the reference or default bandwidth (**href**) was able to produce KDE in adehabitat.  
Estimation with href is NOT reliable for use on multimodal datasets because it results in over-smoothing of home ranges and the multimodal distribution of locations is typical for most species.  
  
Obtain the KDE home range using href  
 - use href instead of LSCV because the isolated home range blobs look odd  
  
```{r kde using href}
kudhref <- kernelUD(locs_1[, 1],
                    h = "href")

```
  
  
Convert the 95% KDE as vectors  
  
```{r 95 kde href}
homerange_locs_1href <- getverticeshr(kudhref, 
                                      percent = 95)

```
  
  
Plot the vector and points  
```{r plot kde href}
plot(homerange_locs_1href,
     border = 1:3,
     lwd = 6)

plot(locs_1,
     col = as.data.frame(locs_1)[, 1],
     add = T)

```
  
  
This does look slightly better than before.  
It is important to relate this to the ecology of the species, which might determine the home range method used.  
  
Export shapefile  
```{r export kde href}
writePolyShape(homerange_locs_1href,
               "HomeRange/HR_data/95kde_href") 

```
  
  
Core Area: Using 50 KDE as vectors  
  
```{r core kdehref}
homerange_locs_1href_core <- getverticeshr(kudhref,
                                           percent = 50)

```
  
  
```{r plot the core area}
plot(homerange_locs_1href,
     border = 1:3,
     lwd = 6)

plot(locs_1,
     col = as.data.frame(locs_1)[, 1],
     add = T)

plot(homerange_locs_1href_core,
     border = 1:3,
     lwd = 4,
     lty = "dashed",
     add = T)

```
  
Final plot showing locations within the typically associated 95% home range boundaries and core areas for common scale-backed antbird _Willisornis poecilinotus_, black-headed antbird _Percnostola rufifrons_, and rufous-capped anthrush _Formicarius colma_.
  
```{r export core shapefile}
writePolyShape(homerange_locs_1href_core,
               "HomeRange/HR_data/core_50_kde_href")

```




